Estimating the replicability of psychology experiments after an initial failure to replicate

### 2024-06-12 manuscript/manuscript_anon.Rmd
submit collabra

---
title: "Estimating the replicability of psychology experiments after an initial failure to replicate"
author:
  - name: Veronica Boyce
    affiliation: Psych
    footnote:
      - corresp
  - name: Ben Prystawski
    affiliation: Psych
  - name: Adani B. Abutto
    affiliation: Psych
  - name: Emily M. Chen
    affiliation: Psych
  - name: Ziwen Chen
    affiliation: GSB
  - name: Howard Chiu
    affiliation: Edu
  - name: Irmak Ergin
    affiliation: Psych
  - name: Anmol Gupta
    affiliation: Psych
  - name: Chuqi Hu
    affiliation: Symsys
  - name: Bendix Kemmann
    affiliation: Philosophy
  - name: Nastasia Klevak
    affiliation: Psych
  - name: Verity Y. Q. Lua
    affiliation: Psych
  - name: Mateus M. Mazzaferro
    affiliation: Edu
  - name: Khaing Mon
    affiliation: Symsys
  - name: Dan Ogunbamowo
    affiliation: Psych
  - name: Alexander Pereira
    affiliation: [Psych, Philosophy]
  - name: Jordan Troutman
    affiliation: CS
  - name: Sarah Tung
    affiliation: Psych
  - name: Raphael Uricher
    affiliation: Psych
  - name: Michael C. Frank
    affiliation: Psych
address:
  - code: Psych
    address: Department of Psychology, Stanford University
  - code: GSB
    address: Graduate School of Business, Stanford University
  - code: Edu
    address: Graduate School of Education, Stanford University
  - code: Symsys
    address: Symbolic Systems Program, Stanford University
  - code: Philosophy
    address: Department of Philosophy, Stanford University
  - code: CS
    address: Department of Computer Science, Stanford University
footnote:
  - code: corresp
    text: "Corresponding author. Email: vboyce@stanford.edu"
bibliography: ["251rescue.bib"] # Replace with one or more of your own bibtex files. Better BibTeX for Zotero is your friend
csl: apa7.csl # Use any CSL
nocite: |
  @krauss2003, @ngo2019, @todd2016, @jara-ettinger2022, @porter2016, @birch2007, @child2018, @chou2016, @craig2014, @gong2019, @haimovitz2016, @hopkins2016, @paxton2012, @payne2008, @schechtman2010, @tarampi2016, @yeshurun2003
link-citations: TRUE
output:
  bookdown::pdf_document2:
    toc: FALSE
    keep_tex: TRUE
    template: generic_article_template.tex
    #md_extensions: "-autolink_bare_uris"
    number_sections: TRUE
    citation_package: default # Can also be "natbib"
lang: en # Main document language in BCP47 format
geometry: "margin=25mm"
papersize: a4
#linestretch: 2 # for double spacing
endfloat: FALSE # Set to TRUE to turn on latex endfloat package to place figures and tables at end of document
# endfloatoption: # See endfloat documentation for more possibilities
numberlines: FALSE
authblk: TRUE # FALSE = author affiliations in footnotes; TRUE = author affiliations in a block below author names
footnotehyper: FALSE # TRUE will give you enhanced table footnote capabilities. Set to FALSE to be able to use French blocks. Needed due to what appears to be a latex bug.
urlcolor: blue
linkcolor: blue
citecolor: blue
graphics: TRUE # Needed to be able to include images
tables: TRUE # Needed to be able to include tables
# fancyhdr:
#   first:
#     #headleft: "REPORT-NO-XXXX"
#     headright: "Kaplan et al. (2021)"
#     headrulewidth: 0pt
#     #footleft: A left foot
#     footrulewidth: 0pt
#   subsequent:
#     #headleft: "NEXT-PAGE-HEADER-LEFT"
#     headright: "Kaplan et al. (2021)"
#     headrulewidth: 1pt
#     footrulewidth: 0pt
header-includes:
 - \usepackage{tikz}
 - \usetikzlibrary{positioning,chains}
 - \usepackage{setspace}\singlespacing
 - \renewcommand{\textfraction}{0.00}
 - \renewcommand{\topfraction}{1}
 - \renewcommand{\bottomfraction}{1}
 - \renewcommand{\floatpagefraction}{1}
 - \setcounter{topnumber}{3}
 - \setcounter{bottomnumber}{3}
 - \setcounter{totalnumber}{4}
editor_options: 
  markdown: 
    wrap: sentence
---

@@

<!-- Re-replication has not been studied much, so it is uncertain how often re-replications recover the results of an original study. That is, it is not well known how definitive a first replication failure is.  -->
Relatively few studies have estimated re-replication success. 
One estimate comes from @ebersole2020. After the original RP:P study reported of a replication rate around 40% [@openscienceconsortium], @gilbert2016 raised concerns that methodological differences might explain the low replication rate. In 11 of the RP:P studies, the original authors had raised concerns about protocol fidelity prior to data collection.
Of these 11 with concerns, 10 failed to obtain significant results in the RP:P replication. 
Thus, @ebersole2020 re-replicated 10 of these 11 in larger samples, using both the RP:P protocol and a new protocol revised under advice of the original authors or other experts. This study thus tested whether larger sample sizes and/or "better" methodologies would "rescue" the failed RP:P replications.
The result was that 2 of the 10 revised protocols found a significant result (but not the 1/11 that originally was successful).
Even when a significant effect was recovered, the effect size was much diminished [@ebersole2020].
<!-- We could frame @ebersole2020 as an attempt to recover the original effects of the 9 studies with significant original effects and non-significant RP:P effects by a combination of high power and design tweaks. -->
The re-replication success rate was thus 2/9 (22%) for studies with non-significant first replications, under favorable, high-resource circumstances.

<!-- While it is tempting to speculate on the reasons for failed replications, the @gilbert2016 and @ebersole2020 example shows these speculations on causes of replication failure do not always pan out in re-replication outcomes. -->

@@

PSYCH 251 is a graduate-level experimental methods class taught by MCF.
In previous years, students have conducted replication projects [@boyce2023; @hawkins2018].
In Fall 2023, students were offered the option to complete a "rescue" project where they re-replicated one of the unsuccessful replications from a previous year.
Students could also opt to do a normal replication instead.
If they opted for a rescue, they were given the option to participate in a collaborative meta-science project and earn authorship on the current paper based on 1) completing a rescue project, 2) reviewing another student's rescue project for reproducibility, and 3) reviewing and approving the final manuscript. 
We report on the result of `r num_proj` rescue projects that opted to be part of the paper and completed data collection.

A spreadsheet of projects, individual project write-ups (both first replications and rescues), individual project data and analyses for rescue projects, and the analytic code for this paper are all available at https://osf.io/cyk5w/.

@@

Students were required to think of reasons the first replication might not have worked, and address them if they could.
A list of possible reasons and solutions was given to students (https://osf.io/dfwk2).
In general, we encouraged students to add manipulation checks as appropriate, and to adapt their materials to online studies, which may benefit from attention checks and/or shorter duration.
For instance, the rescue of @paxton2012 switched from using the cognitive reflection task (CRT) which many online participants are over-familiar with to using the newer CRT-2 which is less overused but attempts to measure the same construct [@thomson2016].
The rescue of @jara-ettinger2022 discovered that the replication had accidentally used an illustrated version of the stimuli rather than the photographic stimuli used in the original and used the photographic stimuli instead.
@tarampi2016 had participants do a timed navigation task where they wrote L or R to indicate which way to turn at each intersection on a paper map. The replication ported this experiment online and had participants click a drop down menu to select left or right for each turn. The rescue removed the unwieldy drop down menus and opted to have participants press keys on the keyboard to indicate the direction of a turn, which seemed like a more natural interface.
Once experimental designs and analytic plans were approved by class TAs (VB and BP), students pre-registered and ran their samples.

With one exception, samples were collected on Prolific [the rescue of @yeshurun2003 ran in-person on the Stanford student subject pool].
We tried to power studies adequately [with a target of 2.5x original following @simonsohn2015], but due to cost constraints, not all studies were powered at this level.
The rescues had on average `r round(m,2)` times the original sample post-exclusion (median: `r round(mid,2)`, IQR: `r round(low,2)` - `r round(high,2)`, minimum: `r round(lowest, 2)`, maximum: `r round(highest,2)`, see Table \@ref(tab:tab-size) for all sample sizes).
Across the `r num_proj-1` Prolific studies, we spent \$`r pluck(cost,1,1) |> round()`, for an average of \$`r (pluck(cost,1,1)/(num_proj-1)) |> round()` per project.

@@

We followed @boyce2023 in the properties of the studies we measured and how we quantified replication success. 
Each project was rated on the basis of subjective replication success (on a 0--1 scale) by both MCF and one of VB and BP.
We thought about replication in terms of general match in key study results (direction, magnitude, and significance), rather than focusing on any singular numeric result or significance cut-off.
Interrater reliability was `r irr |> round(3)`; disagreements were minor (at no point greater than .25) and were resolved through discussion.

@@

<!--Visualizations of the statistical consistency between the original and each replication are shown in TODO FIGURE. The range around the open circle is the 95% predictive interval, representing the interval where, assuming that both are drawn from the same distribution, the replication will fall 95% of the time, given the precision (standard deviation) of the replication [@patil2016]. We again impute a heterogeneity value of $\tau=.21$ SMD for the construction of the prediction intervals.-->

<!--* We will use p-original to evaluate how consistent the original effect size is with the totality of replications. We expect there to be a small number of replications, so we will impute the heterogeneity value as in https://osf.io/preprints/psyarxiv/dpyn6/. -->

@@

An additional factor that influences the interpretation of a replication is how close the replication's methods were to the original.
In the rescue projects, we aimed to have methods be as close as was feasible or appropriate. However, rescue projects varied in how close the re-replications actually were, often due to limitations in the availability of original stimuli and original instructions, in addition to the use of primarily online subject pools.
Table \@ref(tab:tab-size) shows the closeness of each first replication and rescue according to the classification scheme from @lebel2018.
<!-- TODO CLOSENESS NUMBERS SHOULD BE DOUBLE CHECKED AND FIRST REPS REVISED AS NEEDED -->

@@

In cases where re-replications failed, we do not believe any conclusions are licensed about the status of the original results.
Students conducting rescue projects put substantial effort into trying to set up rescues that had a good chance of success, but projects were constrained by budget limitations, a short timeline, and the constraint that most data collection needed to occur online.
These limitations are representative of the sort of resource limitations often faced by early-career researchers.
That said, it is possible that different results might be obtained in better-resourced settings, by scientists with more expertise and more time.
<!-- were false positives, we merely claim the original results do not support cumulative research by early-career researchers under the constrained conditions we tested. -->

@@

@@

We thank the students from previous years of PSYCH 251 who graciously agreed to let us re-replicate their projects and shared their work with us. We thank the other members of the PSYCH 251 teaching team and students for feedback on projects. Data collection was supported by funding from the Stanford Departments of Psychology and Philosophy, Graduate School of Education, and the Symbolic Systems program.

# Author Contributions {.unnumbered}

VB and MCF conceptualized the study. BP contributed project administration, VB contributed formal analysis, and MCF contributed supervision.

VB wrote the first draft, BP and MCF contributed editing, and all authors reviewed the final manuscript.

>>>>

---
title: "Estimating the replicability of psychology experiments after an initial failure to replicate"
author:
  - name: anonymous
bibliography: ["251rescue.bib"] # Replace with one or more of your own bibtex files. Better BibTeX for Zotero is your friend
csl: apa7.csl # Use any CSL
nocite: |
  @krauss2003, @ngo2019, @todd2016, @jara-ettinger2022, @porter2016, @birch2007, @child2018, @chou2016, @craig2014, @gong2019, @haimovitz2016, @hopkins2016, @paxton2012, @payne2008, @schechtman2010, @tarampi2016, @yeshurun2003
link-citations: TRUE
output:
  bookdown::pdf_document2:
    toc: FALSE
    keep_tex: TRUE
    template: generic_article_template.tex
    #md_extensions: "-autolink_bare_uris"
    number_sections: TRUE
    citation_package: default # Can also be "natbib"
lang: en # Main document language in BCP47 format
geometry: "margin=25mm"
papersize: a4
#linestretch: 2 # for double spacing
endfloat: FALSE # Set to TRUE to turn on latex endfloat package to place figures and tables at end of document
# endfloatoption: # See endfloat documentation for more possibilities
numberlines: FALSE
authblk: TRUE # FALSE = author affiliations in footnotes; TRUE = author affiliations in a block below author names
footnotehyper: FALSE # TRUE will give you enhanced table footnote capabilities. Set to FALSE to be able to use French blocks. Needed due to what appears to be a latex bug.
urlcolor: blue
linkcolor: blue
citecolor: blue
graphics: TRUE # Needed to be able to include images
tables: TRUE # Needed to be able to include tables
# fancyhdr:
#   first:
#     #headleft: "REPORT-NO-XXXX"
#     headright: "Kaplan et al. (2021)"
#     headrulewidth: 0pt
#     #footleft: A left foot
#     footrulewidth: 0pt
#   subsequent:
#     #headleft: "NEXT-PAGE-HEADER-LEFT"
#     headright: "Kaplan et al. (2021)"
#     headrulewidth: 1pt
#     footrulewidth: 0pt
header-includes:
 - \usepackage{tikz}
 - \usetikzlibrary{positioning,chains}
 - \usepackage{setspace}\singlespacing
 - \renewcommand{\textfraction}{0.00}
 - \renewcommand{\topfraction}{1}
 - \renewcommand{\bottomfraction}{1}
 - \renewcommand{\floatpagefraction}{1}
 - \setcounter{topnumber}{3}
 - \setcounter{bottomnumber}{3}
 - \setcounter{totalnumber}{4}
editor_options: 
  markdown: 
    wrap: sentence
---

@@

<!-- Re-replication has not been studied much, so it is uncertain how often re-replications recover the results of an original study. That is, it is not well known how definitive a first replication failure is.  -->
Relatively few studies have estimated re-replication success. 
One estimate comes from @ebersole2020. After the original RP:P study reported of a replication rate around 40% [@openscienceconsortium], @gilbert2016 raised concerns that methodological differences might explain the low replication rate. In 11 of the RP:P studies, the original authors had raised concerns about protocol fidelity prior to data collection.
Of these 11 with concerns, 10 failed to obtain significant results in the RP:P replication. 
Thus, @ebersole2020 re-replicated 10 of these 11 in larger samples, using both the RP:P protocol and a new protocol revised under advice of the original authors or other experts. This study thus tested whether larger sample sizes and/or "better" methodologies would "rescue" the failed RP:P replications.
The result was that 2 of the 10 revised protocols found a significant result (but not the 1/11 that originally was successful).
Even when a significant effect was recovered, the effect size was much diminished [@ebersole2020].
The re-replication success rate was thus 2/9 (22%) for studies with non-significant first replications, under favorable, high-resource circumstances.

@@

[class name] is a graduate-level experimental methods class taught by [anonymous].
In previous years, students have conducted replication projects [@boyce2023; @hawkins2018].
In Fall 2023, students were offered the option to complete a "rescue" project where they re-replicated one of the unsuccessful replications from a previous year.
Students could also opt to do a normal replication instead.
If they opted for a rescue, they were given the option to participate in a collaborative meta-science project and earn authorship on the current paper based on 1) completing a rescue project, 2) reviewing another student's rescue project for reproducibility, and 3) reviewing and approving the final manuscript. 
We report on the result of `r num_proj` rescue projects that opted to be part of the paper and completed data collection.

A spreadsheet of projects, individual project write-ups (both first replications and rescues), individual project data and analyses for rescue projects, and the analytic code for this paper are all available at https://osf.io/cyk5w/?view_only=620212fb6b55490d9ae3a555e7b58eca.

@@

Students were required to think of reasons the first replication might not have worked, and address them if they could.
A list of possible reasons and solutions was given to students (https://osf.io/dfwk2).
In general, we encouraged students to add manipulation checks as appropriate, and to adapt their materials to online studies, which may benefit from attention checks and/or shorter duration.
For instance, the rescue of @paxton2012 switched from using the cognitive reflection task (CRT) which many online participants are over-familiar with to using the newer CRT-2 which is less overused but attempts to measure the same construct [@thomson2016].
The rescue of @jara-ettinger2022 discovered that the replication had accidentally used an illustrated version of the stimuli rather than the photographic stimuli used in the original and used the photographic stimuli instead.
@tarampi2016 had participants do a timed navigation task where they wrote L or R to indicate which way to turn at each intersection on a paper map. The replication ported this experiment online and had participants click a drop down menu to select left or right for each turn. The rescue removed the unwieldy drop down menus and opted to have participants press keys on the keyboard to indicate the direction of a turn, which seemed like a more natural interface.
Once experimental designs and analytic plans were approved by class TAs [anonymized], students pre-registered and ran their samples.

With one exception, samples were collected on Prolific [the rescue of @yeshurun2003 ran in-person on the [anonymized university] student subject pool].
We tried to power studies adequately [with a target of 2.5x original following @simonsohn2015], but due to cost constraints, not all studies were powered at this level.
The rescues had on average `r round(m,2)` times the original sample post-exclusion (median: `r round(mid,2)`, IQR: `r round(low,2)` - `r round(high,2)`, minimum: `r round(lowest, 2)`, maximum: `r round(highest,2)`, see Table \@ref(tab:tab-size) for all sample sizes).
Across the `r num_proj-1` Prolific studies, we spent \$`r pluck(cost,1,1) |> round()`, for an average of \$`r (pluck(cost,1,1)/(num_proj-1)) |> round()` per project.

@@

We followed @boyce2023 in the properties of the studies we measured and how we quantified replication success. 
Each project was rated on the basis of subjective replication success (on a 0--1 scale) by both [anonymous] and one of [also anonymous].
We thought about replication in terms of general match in key study results (direction, magnitude, and significance), rather than focusing on any singular numeric result or significance cut-off.
Interrater reliability was `r irr |> round(3)`; disagreements were minor (at no point greater than .25) and were resolved through discussion.

@@

@@

An additional factor that influences the interpretation of a replication is how close the replication's methods were to the original.
In the rescue projects, we aimed to have methods be as close as was feasible or appropriate. However, rescue projects varied in how close the re-replications actually were, often due to limitations in the availability of original stimuli and original instructions, in addition to the use of primarily online subject pools.
Table \@ref(tab:tab-size) shows the closeness of each first replication and rescue according to the classification scheme from @lebel2018.

@@

In cases where re-replications failed, we do not believe any conclusions are licensed about the status of the original results.
Students conducting rescue projects put substantial effort into trying to set up rescues that had a good chance of success, but projects were constrained by budget limitations, a short timeline, and the constraint that most data collection needed to occur online.
These limitations are representative of the sort of resource limitations often faced by early-career researchers.
That said, it is possible that different results might be obtained in better-resourced settings, by scientists with more expertise and more time.

@@

# Contributions {.unnumbered}

- Contributed to conception and design: VB and MCF
- Contributed to acquisition of data: all authors except VB, BP, and MCF
- Contributed to analysis and interpretation of data: all authors
- Drafted and/or revised the article: VB, BP, and MCF
- Approved the submitted version for publication: all authors

<!--VB and MCF conceptualized the study. BP contributed project administration, VB contributed formal analysis, and MCF contributed supervision.

VB wrote the first draft, BP and MCF contributed editing, and all authors reviewed the final manuscript.-->

@@

We thank the students from previous years of PSYCH 251 who graciously agreed to let us re-replicate their projects and shared their work with us. We thank the other members of the PSYCH 251 teaching team and students for feedback on projects.

# Funding {.unnumbered}
Data collection was supported by funding from the Stanford Departments of Psychology and Philosophy, Graduate School of Education, and the Symbolic Systems program.

# Competing Interests {.unnumbered}
The authors declare no competing interests.

# Data accessibility {.unnumbered}

A spreadsheet of projects, individual project write-ups (both first replications and rescues), individual project data and analyses for rescue projects, and the analytic code for this paper are all available at https://osf.io/cyk5w.
